---
title: "Functions"
output: 
 learnr::tutorial:
    progressive: true
    allow_skip: true
    css:
      - https://github.research.chop.edu/pages/CQI/chop-bootstrap/bootstrap-3/bootstrap.min.css
      - https://github.research.chop.edu/pages/CQI/flexdashboard-theme/css/flexdashboard.min.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(data.table)
library(dplyr)
library(rcompanion)
library(FSA)
options(digits = 3, scipen = 9999)
tutorial_options(exercise.completion = TRUE)
knitr::opts_chunk$set(echo = FALSE)
function_name <- function(arg1, arg2) {
	arg1 + arg2
}
my_mean <- function(my_vector) {
      mean(my_vector)
}
my_vector <- c(4, 5, 10)
increment <- function(number, by = 1) {
      number + by
}
remainder <- function(num, divisor = 2) {
      num %% divisor
}
telegram <- function(...){
  paste("START", ..., "STOP")
}
add_two_numbers <- function(num1, num2){
   num1 + num2
}

multiply_two_numbers <- function(num1, num2){
num1 * num2
}

some_function <- function(func){
   func(2, 4)
}
evaluate <- function(func, dat){
  func(dat)
}
boring_function <- function(x) {
	x
}
```

**Hint**: Use the Solution buttons to check your work or to get help if you can't figure out what the solution is.     
**Another Hint**: If you see a red ex next to your code, DO NOT RUN IT or this tutorial might freeze or crash. If it does freeze or crash, simply reload it. Your prior work will have been saved.

## Functions

Functions are fundamental building blocks of the R language. They are small pieces of reusable code that can be treated like any other R object.

If you've worked through any other part of this course, you've probably used some functions already. Functions are usually characterized by the name of the function followed by parentheses. Let's try using a few basic functions just for fun. The `Sys.Date` function returns a string representing today's date. Type `Sys.Date()` below and see what happens.

```{r a, exercise=TRUE}

``` 
```{r a-solution}
Sys.Date()
```

The `mean` function takes a vector of numbers as input, and returns the average of all of the numbers in the input vector. Inputs to functions are often called arguments. Providing arguments to a function is also sometimes called "passing" arguments to that function. Arguments you want to pass to a function go inside the function's parentheses. Try passing the argument `c(2, 4, 5)` to the `mean` function.

```{r c, exercise=TRUE}

```
```{r c-solution}
mean(c(2, 4, 5))
```

Functions usually take arguments which are variables that the function operates on. For example, the `mean` function takes a vector as an argument, like in the case of `mean(c(2,6,8))`. The `mean` function then adds up all of the numbers in the vector and divides that sum by the length of the vector.

## Write your own Function

You're about to write your first function! Just like you would assign a value to a variable with the assignment operator, you assign functions in the following way:

```{r d}
function_name <- function(arg1, arg2) {
	#' Manipulate arguments in some way, for example
	arg1 + arg2
	#' Return a value. Sometimes, as here, manipulate the value and return the argument at
	#' same time. R returns the last thing. You could also do it like this:
	x <- arg1 + arg2 # Manipulate
	x # Return
}
```

The `function_name` you assign will become the name of your function. `arg1` and `arg2` represent the arguments of your function. You can manipulate the arguments you specify within the function. After "sourcing" the function (reading it into memory), you can use it by typing its name and passing anything it needs between the parentheses.

Assign 3.14 to `arg1` and 4 to `arg2`. Then pass both `arg1` and `arg2` to `function_name`.


```{r e, exercise=TRUE}


```
```{r e-solution}
arg1 <- 3.14
arg2 <- 4
function_name(arg1, arg2) 
```

Below we will create a function called `boring_function`. This function takes the argument `x` as input, and returns the value of `x` without modifying it. Make the function workable (by deleting the hashtag in front of the `x`), then run the function by passing it a value. 

Remember, the last R expression to be evaluated in a function always becomes the return value of that function. We want this function to take one argument, `x`, and return `x` without modifying it. Delete the pound sign so that `x` is returned without any modification. Make sure you run the code---but you won't see anything happen as a result. Running a function's code puts the function in memory so you can use it with a function call. 

```{r f, exercise=TRUE}
boring_function <- function(x) {
	#x
}
```

Now that you've created your first function, let's test it. Type `boring_function('My first function!')`. If your function works, it should just return the string: 'My first function!'

```{r h, exercise=TRUE}

```
```{r h-solution}
boring_function('My first function!')
```

Congratulations on writing your first function!

## Using Functions to Gain Insight into how R Works

As John Chambers, the creator of R, once said,

>To understand computations in R, two slogans are helpful:     
1. Everything that exists is an object.    
2. Everything that happens is a function call.

If you want to see the source code for any function, just type the function name without any arguments or parentheses. Let's try this out with the function you just created. Type `boring_function` to view its source code.


```{r k, exercise=TRUE}

```
```{r k-solution}
boring_function
```

Time to make a more useful function. We're going to replicate the functionality of the `mean` function by creating a function called `my_mean`. Remember that to calculate the average of all of the numbers in a vector you find the sum of all the numbers in the vector, and then divide that sum by the number of numbers in the vector.

You're free to implement the function `my_mean` however you want, as long as it returns the average of all of the numbers in `my_vector`. 

>Hint #1: `sum()` returns the sum of a vector.   
	Ex: `sum(c(1, 2, 3))` evaluates to 6

>Hint #2: `length()` returns the size of a vector.    
	Ex: `length(c(1, 2, 3))` evaluates to 3

>Hint #3: The mean of all the numbers in a vector is equal to the sum of all of the numbers in the vector divided by the size of the vector. 

Once you've written your function, test out your `my_mean` function by finding the mean of the vector `c(4, 5, 10)`.

```{r m, exercise=TRUE, , exercise.lines = 15}
my_vector <- c(4, 5, 10)
# Write your function here to find the mean of my_vector:



```
```{r m-solution}
my_vector <- c(4, 5, 10)
my_mean <- function(x) {
  mean(x)
}
my_mean(c(4, 5, 10))
```

Next, let's try writing a function with default arguments. You can set default values for a function's arguments, and this can be useful if you think someone who uses your function will set a certain argument to the same value most of the time.

Let me show you an example of a function I'm going to make up called `increment`. Most of the time I want to use this function to increase the value of a number by one. This function will take two arguments: `number` and `by`, where `number` is the digit I want to increment and `by` is the amount I want to increment `number` by. I've written the function below. 

```{r n, echo = TRUE}
increment <- function(number, by = 1) {
      number + by
}
```

If you take a look in between the parentheses you can see that I've set `by` equal to 1. This means that the `by` argument will have the default value of 1. 

I can now use the `increment` function without providing a value for `by`: `increment(5)` will evaluate to 6. 

However if I want to provide a value for the `by` argument I still can! The expression: `increment(5, 2)` will evaluate to 7. 

You're going to write a function called `remainder`. `remainder` will take two arguments: `num` and `divisor`, where `num` is divided by `divisor` and the `remainder` is returned. Imagine that you usually want to know the remainder when you divide by 2, so set the default value of `divisor` to 2. Please be sure that `num` is the first argument and `divisor` is the second argument.

>Hint #1: You can use the modulus operator %% to find the remainder.    
Ex: 7 %% 4 evaluates to 3. 

Remember to set appropriate default values! After you have written the function, type `remainder(5)` to see what happens.

```{r o, exercise=TRUE}
 # Write your code here.
 # Remember: the last expression evaluated will be returned.
```
```{r o-solution}
remainder <- function(num, divisor = 2) {
      num %% divisor
}
remainder(5)
```

If the answer you got was 1, then you probably wrote good code. 

You provided one argument to the function, and R matched that argument to `num` since `num` is the first argument. The default value for `divisor` is 2, so the function used the default value you provided.

Now let's test the `remainder` function by providing two arguments. 

Copy your function from above into the last code box and paste it into the next one. Unfortunately, we need you to do this because it won't be remembered between code boxes. Lessons like the one you're doing right now are based upon the R package `learnr`, and `learnr` does not remember variables or functions from one code box to the next. We are working on improving this, but in the meantime, we appreciate your willingness to do all this copying and pasting so this interactive lesson works as it is supposed to. 

After you paste the function into the code box, add `remainder(11, 5)`, run the code, and let's see what happens. 

```{r qq, exercise=TRUE}

```
```{r qq-solution}
remainder <- function(num, divisor = 2) {
      num %% divisor
}
remainder(11,5)
```

Once again, the arguments have been matched appropriately if the answer was 1. 

You can also explicitly specify arguments in a function. When you explicitly designate argument values by name, the ordering of the arguments becomes unimportant. Copy your function definition again, then try this out by typing `remainder(divisor = 11, num = 5)`.

```{r s, exercise=TRUE}

```
```{r s-solution}
remainder <- function(num, divisor = 2) {
      num %% divisor
}
remainder(divisor = 11, num = 5)
```
```{r s-check}
remainder(divisor = 11, num = 5)
```

As you can see, there is a significant difference between `remainder(11, 5)` and `remainder(divisor = 11, num = 5)`! The answer this time should be 5.

R can also partially match arguments. Try typing `remainder(4, div = 2)` to see this feature in action. Copy your function definition again first. 

```{r r, exercise=TRUE}

```
```{r r-solution}
remainder <- function(num, divisor = 2) {
      num %% divisor
}
remainder(4, div = 2)
```

If you got 0, then your code has now been thoroughly tested and works as required. If you didn't get 0, then you might want to fix your code.

A word of warning: in general you want to make your code as easy to understand as possible. Switching around the orders of arguments by specifying their names or only using partial argument names can be confusing, so use these features with caution.

With all of this talk about arguments, you may be wondering if there is a way you can see a function's arguments (besides looking at the documentation). Thankfully, you can use the `args` function. 

Again bringing your function definition along, type `args(remainder)` to examine the arguments for the `remainder` function.

```{r q, exercise=TRUE}

```
```{r q-solution}
remainder <- function(num, divisor = 2) {
      num %% divisor
}
args(remainder)
```

`function (num, divisor = 2)` means that the function `function` has two arguments, `num` and `divisor`, which you set to have a default value of 2. 

You may not realize it, but I just tricked you into doing something pretty interesting. `args()` is a function, `remainder()` is a function, yet `remainder` was an argument for `args()`. Yes, it's true: you can pass functions as arguments! This is a very powerful concept. Let's write a script to see how it works.

## Functions as Arguments

You can pass functions as arguments to other functions just like you can pass data to functions. Let's say you define the following functions:

```{r newfunctions, echo = TRUE}
add_two_numbers <- function(num1, num2){
   num1 + num2
}

multiply_two_numbers <- function(num1, num2){
num1 * num2
}

some_function <- function(func){
   func(2, 4)
}

some_function(add_two_numbers)

some_function(multiply_two_numbers)
```

As you can see, we used the argument name `func` like a function inside of `some_function()`. By passing functions as arguments, `some_function(add_two_numbers)` evaluated to 6, while `some_function(multiply_two_numbers)` evaluated to 8.

Finish the function definition below so that if a function is passed into the `func` argument and some data (like a vector) is passed into the `dat` argument, `evaluate` will return the result of `dat` being passed as an argument to `func`.

>Hints: This exercise is a little tricky so I'll provide a few example of how evaluate() should act:    
   1. evaluate(sum, c(2, 4, 6)) should evaluate to 12    
   2. evaluate(median, c(7, 40, 9)) should evaluate to 9    
   3. evaluate(floor, 11.1) should evaluate to 11
   

```{r t, exercise=TRUE}

```
```{r t-solution}
evaluate <- function(func, dat){
  func(dat)
}
```

Let's take your new `evaluate()` function for a spin. Use `evaluate` to find the standard deviation of the vector `c(1.4, 3.6, 7.9, 8.8)`. As usual, copy your function into the code box, then construct your call to it. 
   
```{r ss, exercise=TRUE}

```
```{r ss-solution}
evaluate <- function(func, dat){
  func(dat)
}
evaluate(sd, c(1.4, 3.6, 7.9, 8.8))
```

If you did not get 3.51 as an answer, have another look at your code or at your call to it. Many problems with code functionality are typos. 

The idea of passing functions as arguments to other functions is an important and fundamental concept in programming. You may be surprised to learn that you can pass a function as an argument without first defining the passed function. Functions that are not named are appropriately known as _anonymous functions_.

## Anonymous Functions

Let's use the `evaluate` function to explore how anonymous functions work. For the first argument of `evaluate`, we're going to write a tiny function that fits on one line. In the second argument we'll pass some data to the tiny anonymous function in the first argument.

Paste your `evaluate` function in the code box, then type the following command and then we'll discuss how it works: `evaluate(function(x) {x + 1}, 6)`.

```{r u, exercise = TRUE}

```
```{r u-solution}
evaluate <- function(func, dat){
  func(dat)
}
evaluate(function(x) {x + 1}, 6)
```

The first argument is a tiny anonymous function that takes one argument `x` and returns `x + 1`. We passed the number 6 into this function so the entire expression evaluates to 7.

Try using `evaluate` and pass it an anonymous function to return the first element of the vector `c(8, 4, 0)`. Your anonymous function should take only one argument, which should be a variable `x`.

```{r v, exercise = TRUE}

```
```{r v-solution}
evaluate <- function(func, dat){
  func(dat)
}
evaluate(function(x) {x[1]}, c(8, 4, 0))
```

Now try using `evaluate()` along with an anonymous function to return the last element of the vector `c(8, 4, 0)`. Your anonymous function should take only one argument, which should be a variable `x`.

```{r cw, exercise = TRUE}

```
```{r cw-solution}
evaluate <- function(func, dat){
  func(dat)
}
evaluate(function(x) {x[length(x)]}, c(8, 4, 0))
```

That was a tricky one with many possible solutions. If you got 0 as a result, well done!

## Ellipses and paste()

For the rest of the course, we're going to use the `paste()` function frequently. Type `?paste` so we can take a look at the documentation.

```{r aw, exercise = TRUE}

```
```{r aw-solution}
?paste
```

As you can see, the first argument of `paste()` is `...`, which is an ellipsis. The ellipsis allows an indefinite number of arguments to be passed into a function. In the case of `paste()`, any number of strings can be passed as arguments and `paste()` will return all of the strings combined into one string.

Just to see how `paste()` works, type `paste("Programming", "is", "fun!")`

```{r w, exercise = TRUE}

```
```{r w-solution}
paste("Programming", "is", "fun!")
```

An ellipsis can be used to pass arguments to other functions that are used _within_ the function you're writing. Usually a function that has an ellipsis as an argument has the ellipsis as the last argument. The usage of such a function would look like this:

>ellipsis_func(arg1, arg2 = TRUE, ...)

In the above example, `arg1` has no default value, so a value must be provided for it. `arg2` has a default value, and other arguments can come after `arg2` depending on how they're defined in the ellipsis_func() documentation. Interestingly, the usage for the paste function is as follows:

>paste (..., sep = " ", collapse = NULL)

Notice that the ellipsis is the first argument, and all other arguments after the ellipsis have default values. This is a strict rule in R programming: all arguments after an ellipsis must have default values. Take a look at the `simon_says` function below:

```{r simonsays, echo = TRUE}
simon_says <- function(...){    
  paste("Simon says:", ...)    
}
```

`simon_says` works just like the `paste` function, except the begining of every string is prepended by the string "Simon says:".

Telegrams used to be peppered with the words START and STOP in order to demarcate the beginning and end of sentences. Write a function below called `telegram` that formats sentences for telegrams. For example, the expression `telegram("Good", "morning")` should evaluate to `"START Good morning STOP"`. Then test out your telegram function. Use your new telegram function passing in whatever arguments you wish!

```{r aww, exercise = TRUE}

```
```{r aww-solution}
telegram <- function(...){
  paste("START", ..., "STOP")
}
telegram("I am an argument that was wanted here")
```

## Mad Libs

Have you ever played Mad Libs before? In the function below, we will construct a sentence from parts of speech that you provide as arguments. We'll write most of the function, but you'll need to unpack the appropriate arguments from the
ellipsis.

```{r y, exercise=TRUE}
mad_libs <- function(...){
  # Do your argument unpacking here!
  
  # Don't modify any code below this comment.
  # Notice the variables you'll need to create in order for the code below to
  # be functional!
  paste("News from", place, "today where", adjective, "students took to the streets in protest of the new", noun, "being installed on campus.")
}
```
```{r y-solution}
mad_libs <- function(...){
  # Do your argument unpacking here!
  args <- list(...)
  place <- args[["place"]]
  adjective <- args[["adjective"]]
  noun <- args[["noun"]]
  # Don't modify any code below this comment.
  # Notice the variables you'll need to create in order for the code below to
  # be functional!
  paste("News from", place, "today where", adjective, "students took to the streets in protest of the new", noun, "being installed on campus.")
}
```

Equally important to your function code is your calling syntax. This is how you call the function you just created.

```{r prepare-z}
mad_libs <- function(...){
  # Do your argument unpacking here!
  args <- list(...)
  place <- args[["place"]]
  adjective <- args[["adjective"]]
  noun <- args[["noun"]]
  # Don't modify any code below this comment.
  # Notice the variables you'll need to create in order for the code below to
  # be functional!
  paste("News from", place, "today where", adjective, "students took to the streets in protest of the new", noun, "being installed on campus.")
}
```
```{r z, exercise=TRUE}

```
```{r z-solution}
mad_libs(place = "Philadelphia", adjective = "hyper", noun = "dishwasher")
```

Note that any values passed after being assigned to a name using `<-` instead of `=` will fail to appear in the function's output. You can experiment with that above. 

## Your Very Own Binary Operators

The syntax for creating new binary operators is unlike anything else in R. It is syntax for defining new syntax! I would recommend making your own binary operator only if you plan on using it often.

User-defined binary operators have the following syntax:

>     `%[whatever]%`    
where `[whatever]` represents any valid variable name.

Let's say I wanted to define a binary operator that multiplied two numbers and then added one to the product. An implementation of that operator is below:

```{r}
"%mult_add_one%" <- function(left, right) { # Notice the quotation marks!    
  left * right + 1
}
```

I could then use this binary operator like `4 %mult_add_one% 5`, which would evaluate to 21.

Write your own binary operator below from absolute scratch. Your binary operator must be called `%p%` so that the expression

>"Good" %p% "job!"

will evaluate to: `"Good job!"`

```{r aa, exercise=TRUE}

```
```{r aa-solution}
"%p%" <- function(left, right){ 
  paste(left, right)
}
```

You made your own binary operator! Let's test it out. Paste together the strings: 'I', 'love', 'R!' using your new binary operator.

```{r prepare-ab}
"%p%" <- function(left, right){ 
  paste(left, right)
}
```
```{r ab, exercise=TRUE}

```
```{r ab-solution}
"I" %p% "love" %p% "R!"
```

You may wonder when you'll ever want to create your own binary operators. Some people use them for complex math like adding together directions for quarks, and others use them for matrix math. 

Now that you understand how to create your own functions, you can create R in your own image if you like. Start thinking about the packages _you_ want to write!