---
title: "lm output under a microscope"
author: "Sheila Braun, CHOP"
date: "10/18/2019"
output: html_document
---

### `coefficients`

What happens when you look at `egg_fit$coefficients`?

```{r egg_fit-coefficients, exercise = TRUE, exercise.lines = 8}

```
```{r egg_fit-coefficients-solution}
egg_fit$coefficients
```

There we find the information related to our best-fit line. We now know that the line is $\displaystyle Eggs = 12.7 + 1.6 * Weight$. If we know `Weight`, we can predict `Eggs`. To a certain degree. Let's look at the rest of the output. 

### `effects`

We are now using `ln_egg_fit` instead of `egg_fit`. If you don't know why, scroll up a little. 

```{r log-egg-fit-effects, exercise = TRUE, exercise.lines = 8}

```
```{r log-egg-fit-effects-solution}
ln_egg_fit$effects
```

`effects` is an $\displaystyle n$-long vector of orthogonal single-degree-of-freedom effects. These are computed using the formula $\displaystyle t(Q) %*% Y$: Transpose the matrix, then multiply it by $\displaystyle Y$. `%*%` is a multiplication operator for a matrix. The first rank of effects correspond to non-aliased coeffcients (coefficients without collinearity). 

Here's a sandbox so you can play with that concept a bit using the data in this lesson. 

```{r sandbox-3, exercise = TRUE, exercise.lines = 15}

```
```{r sandbox-3-solution}

```

### `rank`


