---
title: "Introduction to the tidyverse: dplyr"
output: 
 learnr::tutorial:
      progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
options(digits = 3, scipen = 9999)
if(!require(learnr)){install.packages("learnr")}
library(learnr)
library(dplyr)
knitr::opts_chunk$set(echo = FALSE)
options(digits = 3, scipen = 9999)
tutorial_options(exercise.completion = FALSE,
                 exercise.eval = TRUE,
                 exercise.lines = 8,
                 exercise.diagnostics = TRUE)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv"
msleep_fname <- basename(url)
msleep <- read.csv(url)
head(msleep)
sleepData <- select(msleep, name, sleep_total)

```

Tip: To check your work, or to see our code for a code chunk that you just can't (or don't have time to) figure out, click on its Solution button.

## What is `dplyr`?

`dplyr` is an R package you can use to transform and summarize tabular data---that is, data with rows and columns. 

To learn more about `dplyr`, see the package vignette [Introduction to dplyr](https://rdrr.io/cran/dplyr/f/vignettes/dplyr.Rmd).

### Why is `dplyr` Useful?

`dplyr` contains a set of functions (or "verbs") that perform common data manipulation operations. For example, `dplyr` is useful for filtering rows, selecting specific columns, re-ordering rows or columns, adding new columns, and summarizing data. 

In addition, `dplyr` contains a useful function to perform another common task,  "split-apply-combine": more about that later. 

### How does `dplyr` Compare to Using Base Functions in R?

If you are familiar with R, you are probably familiar with base R functions such as `split`, `subset`, `apply`, `sapply`, `lapply`, `tapply` and `aggregate`. Some people find the syntax in the `dplyr` package to be easier to work with. I personally like to use Base R for some things and `dplyr` for others. You'll develop your own preferences as you progress in your understanding of R and its many packages. 

## How do I get `dplyr`? 

To install `dplyr` use this syntax (which is typical for installing packages in R): `install.packages("dplyr")`. You can't install any packages because this lesson doesn't run on your computer (it's in the cloud, as you may have noticed), but you can copy the following code to your own RStudio on your own computer:

`install.packages("dplyr")`

To load `dplyr` into the library, type `library(dplyr)`. You don't need quotation marks this time. You can try it here:

```{r b, exercise = TRUE, exercise.lines = 8}

```
```{r b-solution}
library(dplyr)
```

## Mammals Sleep Data

The `msleep` (mammals sleep) data set contains the sleep times and weights for a set of mammals and is available in the dagdata repository on GitHub. This data set contains 83 rows and 11 variables.  

Download the `msleep` data set in CSV format from [here](https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv), and then load into R. I've provided the code for you. Note that there is an interesting function, `basename`, that extracts the file name from a url. We're storing that in `msleep_filename`.  Add `head(msleep)` to the bottom of the code in the code chunk here, then run the code to see the top of the `msleep` data. 

```{r c, exercise = TRUE, exercise.lines = 10}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv"
msleep_fname <- basename(url)
msleep <- read.csv(url)
#Your code:

```
```{r c-solution}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv"
msleep_fname <- basename(url)
msleep <- read.csv(url)
head(msleep)
```

The columns (in order) correspond to the following: 

Column Name | Description
--- | ---
`name` | common name
`genus` | taxonomic rank
`vore` | carnivore, omnivore or herbivore?
`order` | taxonomic rank
`conservation` | the conservation status of the mammal
`sleep_total` | total amount of sleep, in hours
`sleep_rem` | rem sleep, in hours
`sleep_cycle` | length of sleep cycle, in hours
`awake` | amount of time spent awake, in hours
`brainwt` | brain weight in kilograms
`bodywt` | body weight in kilograms

## `dplyr` Verbs

`dplyr` thinks of its function calls as verbs, and here they are:

dplyr verbs | Description
--- | ---
`select` | select columns 
`filter` | filter rows
`arrange` | re-order or arrange rows
`mutate` | create new columns
`summarise` | summarise values
`group_by` | allows for group operations in the "split-apply-combine" concept

The next sections will provide you with practice using `dplyr` verbs. We'll start with the two most basic functions, `select` (for choosing columns) and `filter` (for choosing rows). 

```{r tf-verbs-functions, echo=FALSE}
question("True or false: A verb by any other name is still a function call.",
         answer("True", correct = TRUE),
         answer("False"),
         random_answer_order = TRUE,
         allow_retry = TRUE,
         correct = "You are amazing!",
         incorrect = "Actually, it is."
)
```

## Choosing Columns using `select`

Choose a set of columns by simply listing them in the call to `select` along with the data set name. Put the data set name first, then add the names of the columns you want to choose. Try that now: choose `name` and `sleep_total` from the `msleep` data set and assign them to the variable `sleepData`. Then take a look at the head of `sleepData`. 

```{r d, exercise = TRUE, exercise.lines = 8}

```
```{r d-solution}
sleepData <- select(msleep, name, sleep_total)
head(sleepData)
```

To choose all the columns *except* a specific column, use the "-" (subtraction) operator (also known as negative indexing). Do this now with the `msleep` data set: take everything except `name`. Pass the call to `select` to `head` in a nested fashion so you see the first six lines of the results. 

```{r e, exercise = TRUE, exercise.lines = 8}

```
```{r e-solution}
head(select(msleep, -name))
```

Now select a range of columns by name, use the ":" (colon) operator. We want to see the first six lines of `msleep` for the columns from `name` to `order`.

```{r f, exercise = TRUE, exercise.lines = 8}

```
```{r f-solution}
head(select(msleep, name:order))
```

To select all columns that start with the character string "sl", use the function `starts_with`. Nest a call to `starts_with` inside a call to `select` inside a call to `head`. We want all the columns that start with "sl". 

```{r g, exercise = TRUE, exercise.lines = 8}

```
```{r g-solution}
head(select(msleep, starts_with("sl")))
```

Some additional options to select columns based on a specific criteria include

1. `ends_with` = Select columns that end with a character string
2. `contains` = Select columns that contain a character string
3. `matches` = Select columns that match a regular expression
4. `one_of` = Select column names that are from a group of names

## Choosing Rows using `filter`

Filter the rows for mammals that sleep a total of more than 16 hours. Use the `filter` function and pass it the data set name `msleep` and then a condition the `sleep_total` be greater than or equal to 16. 

You can use the boolean operators (e.g. >, <, >=, <=, !=, %in%) to create the logical tests. If you are not familiar with how to use conditional logic, download and run the [Logic](https://console.cloud.google.com/storage/browser/_details/public-braunsb-media/Arcus-Education-Assets/Lessons/Logic.Rmd?project=arcus-edu-public-c62a737c) lesson for some practice. It's based on the `swirl` lesson of the same name. 

```{r h, exercise = TRUE, exercise.lines = 8}

```
```{r h-solution}
filter(msleep, sleep_total >= 16)
```

Filter the rows for mammals that sleep a total of more than 16 hours *and* have a body weight of greater than 1 kilogram.

```{r i, exercise = TRUE, exercise.lines = 8}

```
```{r i-solution}
filter(msleep, sleep_total >= 16, bodywt >= 1)
```

Filter the rows for mammals in the Perissodactyla and Primates taxonomic order. Use `order %in% c("Perissodactyla", "Primates")` in your syntax.

```{r j, exercise = TRUE, exercise.lines = 8}

```
```{r j-solution}
filter(msleep, order %in% c("Perissodactyla", "Primates"))
```


## Organizing Readable Code using the Pipe Operator: `%>%`

`dplyr` imports the pipe operator (`%>%`)from another package (`magrittr`). This operator allows you to "pipe" the output from one function and use it as the input of another function. Instead of nesting functions, which forces one to read from the inside to the outside, piping allows one to read the functions from left to right or top to bottom in a narrative order. 

Remember `head(select(msleep, name, sleep_total))` from above? Have a go at using piping to send `msleep` to `select`, then to send the input of `select` to `head`. I like to put a newline after each operator, then indent the next line a bit for more readable code. Tabs won't work in your exercise code chunks, so use spaces to indent lines after the first line of a piping sequence. As always, use the Solution button if you get stuck. 

```{r k, exercise = TRUE, exercise.lines = 8}

```
```{r k-solution}
msleep %>% 
    select(name, sleep_total) %>% 
    head
```
```{r u, echo=FALSE}
question("What was your output?",
         answer("The first six lines of `name` and `sleep_total` from `msleep`.", correct = TRUE),
         answer("The first six lines of `msleep`."),
         answer("An `Error` message."),
         random_answer_order = TRUE,
         allow_retry = TRUE
)
```

You will soon see how useful the pipe operator is when we start to combine many functions rather than just a few. Keep in mind what that code would look like _without_ a pipe function.  

Now that you know about the pipe operator (%>%), we will use it throughout the rest of this tutorial. 

## Arranging or Re-Arranging Rows using `arrange`

To `arrange` (or re-order) rows by a particular column, such as the taxonomic `order`, list the name of the column by which you want to arrange the rows. Start by piping `msleep` to `arrange`, and pass `order` to arrange. Then pass the result to `head` so you can see what your code produces. 

```{r m, exercise = TRUE, exercise.lines = 8}

```
```{r m-solution}
msleep %>% arrange(order) %>% head
```

Now let's select three columns, `name`, `order`, and `sleep_total`, from `msleep`, arrange the rows by the taxonomic `order`, then arrange the rows by `sleep_total`. Finally, show the `head` of the final data frame.

```{r n, exercise = TRUE, exercise.lines = 8}

```
```{r n-solution}
msleep %>% 
    select(name, order, sleep_total) %>%
    arrange(order, sleep_total) %>% 
    head
```

Now reuse the code from above, but filter the rows for mammals that sleep for 16 or more hours. There is no need to limit the output to `head` because there are only 8 resulting rows.

```{r o, exercise = TRUE, exercise.lines = 8}

```
```{r o-solution}
msleep %>% 
    select(name, order, sleep_total) %>%
    arrange(order, sleep_total) %>% 
    filter(sleep_total >= 16)
```
```{r w, echo=FALSE}
question("Should the call to `filter` come before or after the call to `select`?",
         answer("Before, because then the data set will be smaller so the `arrange` call will take less time."),
         answer("After, because then the data set will be smaller so the `filter` call will take less time."),
         answer("It depends on your data set and resources", correct = TRUE),
         answer("It doesn't matter"),
         random_answer_order = TRUE,
         allow_retry = TRUE
)
```

Now do the same thing you just did, but `arrange` the rows in the `sleep_total` column in a descending order by wrapping `sleep_total` with the function `desc`. Yes, it's a bit of nesting in a bit of piping; you sometimes pass functions to `arrange` to affect its output. 

```{r p, exercise = TRUE, exercise.lines = 8}

```
```{r p-solution}
msleep %>% 
    select(name, order, sleep_total) %>%
    arrange(order, desc(sleep_total)) %>% 
    filter(sleep_total >= 16)
```

## Creating New Columns using `mutate`

The `mutate` function will add new columns to the data frame or change ones that are already there. Create a new column called `rem_proportion`, which is the ratio of rem sleep to total amount of sleep. Use `names(msleep)` if you have forgotten these variable names. Pipe `msleep` to `mutate`, and in the `mutate` call, set the new column name `rem_proportion` equal to the ratio. Then pipe the result to head. Note that the result isn't being saved anywhere.

```{r q, exercise = TRUE, exercise.lines = 8}

```
```{r q-solution}
msleep %>% 
    mutate(rem_proportion = sleep_rem / sleep_total) %>%
    head
```

You can multiple new columns using a single `mutate` call with what you pass it separated by commas. Try adding more to the `mutate` function call you just constructed: a column called `bodywt_grams` calculated from `bodywt` time one thousand. Add a comma after the input to `mutate` above and place your new code after the comma. Pipe the result to head. 

```{r s, exercise = TRUE, exercise.lines = 8}

```
```{r s-solution}
msleep %>% 
    mutate(rem_proportion = sleep_rem / sleep_total, 
           bodywt_grams = bodywt * 1000) %>%
    head
```

You should see two new columns in the output, `rem_proportion` and `bodywt_grams`. 

## Creating Summaries of the Data Frame using `summarise`

The `summarise` function creates summary statistics, statistics that are aggregated from individual cases, for a given column in the data frame. For instance, summarise can return the mean `sleep_total`. Try that. Call this summary variable `avg_sleep`. Note that you can't just create a new column from summarise because the new column would have a shorter length than the rest of the data frame. What you create from `summarise`, usually, is new variables. 

```{r r, exercise = TRUE, exercise.lines = 8}

```
```{r r-solution}
msleep %>% 
    summarise(avg_sleep = mean(sleep_total))
```

There are many other summary statistics you could consider: `sd`, `min`, `max`, `median`, `sum`, `n` (returns the length of a vector), `first` (returns first value in vector), `last` (returns last value in vector) and `n_distinct` (returns the number of distinct values in vector). 

Try piping `msleep` to summarise, then create three new variables: `avg_sleep` from the mean of `sleep_total`, `min_sleep` from the `min` of `sleep_total`, and `max_sleep` by find `max(sleep_total)`. Then create the variable `total`, equal to the `n` of `sleep_total`. Separate these aggregation commands with commas; wrap them in a single call to `summarise`. 

```{r v, exercise = TRUE, exercise.lines = 8}

```
```{r v-solution}
msleep %>% 
    summarise(avg_sleep = mean(sleep_total), 
              min_sleep = min(sleep_total),
              max_sleep = max(sleep_total),
              total = n())
```

    
## Grouping Results using `group_by`

The `group_by` verb is an important function in `dplyr`. This is where we get back to "split-apply-combine". We literally want to split the data frame by some variable (e.g. taxonomic `order`), apply a function to the individual data frames that result from the split, and then combine the output.   

Let's do that: split the `msleep` data frame by the taxonomic `order`, then ask for the same summary statistics as above. We expect a set of summary statistics for each element of `order`. 

```{r t, exercise = TRUE, exercise.lines = 8}

```
```{r t-solution}
msleep %>% 
    group_by(order) %>%
    summarise(avg_sleep = mean(sleep_total), 
              min_sleep = min(sleep_total), 
              max_sleep = max(sleep_total),
              total = n())
```


## Parting Advice---or Not Parting Advice

Now that you've done this tutorial, I recommend you do the exercises in it again, this time without making use of the Solution button. There is this tiny "Start Over" text below the table of contents. Be brave: click on it and fill things out. You'll do much better the second time around, and the practice will anchor the ideas in your mind, making you a more fluent R user. 