---
title: "ggplot2 Part 2"
output: 
 learnr::tutorial:
    progressive: true
    allow_skip: true
    css:
      - www/bootstrap.min.css
      - www/flexdashboard.min.css
      - www/style.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 9999, digits = 3)
tutorial_options(exercise.completion = FALSE,
                 exercise.eval = TRUE,
                 exercise.lines = 5,
                 exercise.diagnostics = TRUE)

# Variables needed in code chunks below
g <- ggplot(mpg, aes(displ, hwy)) 
set.seed(9699)
x <- rnorm(100, mean = 0, sd = 1)
y <- rnorm(100, mean = 0, sd = 1)
testdat <- cbind(x, y)
testdat <- as.data.frame(testdat)
testdat$y[50] <- 100
myx <- testdat$x
myy <- testdat$y
o <- ggplot(testdat, aes(x = myx, y = myy)) 
ab <- ggplot(mpg, aes(x = displ, y = hwy, color = factor(year)))
```
  
**Hint**: Use the "Hints" and "More hints" buttons to check your work or to get help if you can't figure out what the solution is. The final hint is the solution.      


## Where We've Been: `qplot`

In a previous lesson we showed you the vast capabilities of `qplot`, the basic workhorse function of the `ggplot2` package. In this lesson, we'll focus on some fundamental components of the package. These  underlie `qplot`, which uses default values when it calls them. If you understand these building blocks, you will be better able to customize your plots. 

We'll use the second workhorse function in the package, `ggplot`, as well as other graphing functions. 


```{r ggplot-stands-for, echo=FALSE}
question("Do you remember what the 'gg' of `ggplot2` stands for?",
         answer("goto graphics"),
         answer("grammar of graphics", correct = TRUE),
         answer("good grief"),
         answer("nothing"),
         random_answer_order = TRUE,
         allow_retry = TRUE,
         incorrect = "Think about nouns, verbs, and adjectives.",
         correct = "You are amazing!"
)
```

A "grammar" of graphics means that `ggplot2` contains building blocks with which you can create your own graphical objects. What are these basic building blocks of `ggplot2` plots, you ask? There are 7 of them. 

Obviously, there's a **data frame** which contains the data you're trying to plot. Then the **aesthetic mappings** determine how data is mapped to color, size, etc. The `geoms` (geometric objects) are what you see in the plot (`point`s, `line`s, `shape`s), and `facets` are the panels used in conditional plots. You've used these or seen them used in the first `ggplot2` (`qplot`) lesson.


There are 3 more: `stats` are statistical transformations such as binning, quantiles, and smoothing, which `ggplot2` applies to the data. `scales` show what coding an aesthetic map uses (for example, `male = red`, `female = blue`). Finally, the plots are depicted on a **coordinate system** (e.g., Cartesian). When you used `qplot`, these were taken care of for you.

```{r artists-palette, echo=FALSE}
question("Do you remember what the artist's palette model is in the context of plotting?",
         answer("we draw pictures"),
         answer("we mix paints"),
         answer("plots are built up in layers", correct = TRUE),
         answer("things get messy"),
         random_answer_order = TRUE,
         allow_retry = TRUE, 
         incorrect = "Sounds like fun, but no.",
         correct = "Yes! We build plots up in layers just as an artist creates a good painting."
)
```

As in the `base` plotting system (and in contrast to the `lattice` system), when building plots with `ggplot2`,  the plots are built up in layers, maybe in several steps. You can plot the data, then overlay a summary (for instance, a regression line or smoother) and then add any metadata and annotations you need.

We'll keep using the `mpg` data that comes with the `ggplot2` package. Recall the versatility of `qplot`. Just as a refresher, call `qplot` now with 5 arguments: the first 3 deal with data---`displ`, `hwy`, and `data = mpg`. The fourth is `geom` set equal to the concatenation of the two strings, "point" and "smooth". The fifth is `facets` set equal to the formula `. ~ drv`. Try this now. 

```{r qplot, exercise = TRUE, exercise.lines = 5}

```
```{r qplot-hint-1}
qplot(...)
```
```{r qplot-hint-2}
qplot(displ, hwy, ...)
```
```{r qplot-hint-3}
qplot(displ, hwy, data = mpg, ...)
```
```{r qplot-hint-4}
qplot(displ, hwy, data = mpg, geom = c("point", "smooth"), ...)
```
```{r qplot-solution}
qplot(displ, hwy, data = mpg, geom = c("point", "smooth"), facets = . ~ drv)
```

We see a 3-facet plot, one for each drive type (4, f, and r). `ggplot` also gave us some information about the kind of line it drew and with what formula. 

## Build your First Plot with `ggplot`

First we'll see how `ggplot` works. We'll build up a similar plot using the basic components of the package. We'll do this in a series of steps. 

First we'll create a variable `g` by assigning to it the output of a call to `ggplot` with 2 arguments: the first is `mpg` (our dataset) and the second will tell `ggplot` what we want to plot, in this case, `displ` and `hwy`. These are what we want our aesthetics to represent, so we enclose these as two arguments to the function `aes`. Try this now. 

```{r first-ggplot, exercise = TRUE, exercise.lines = 5}

```
```{r first-ggplot-hint-1}
g <- ggplot(...)
```
```{r first-ggplot-hint-2}
g <- ggplot(mpg, ...)
```
```{r first-ggplot-hint-3}
g <- ggplot(mpg, aes())
```
```{r first-ggplot-hint-4}
g <- ggplot(mpg, aes(displ, ...))
```
```{r first-ggplot-solution}
g <- ggplot(mpg, aes(displ, hwy))
```

Nothing happened, right? As in the `lattice` system, `ggplot` created a *graphical object* which we assigned to the variable `g`. 

Run the R command `summary` with `g` as its argument to see what `g` contains.

```{r summary-g, exercise = TRUE, exercise.lines = 5}

```
```{r summary-g-hint}
summary(...)
```
```{r summary-g-solution}
summary(g)
```

So `g` contains the `mpg` data with all its named components in a 234 by 11 matrix. It also contains a `mapping`, `x` (`displ`), and `y` (`hwy`) which you specified, and no faceting. 

Note that if you tried to print `g` with the expressions `g` or `print(g)` you'd get an error! Even though it's a great package, `ggplot` doesn't know how to display the data yet since, although you specified what data goes _into_ the plot, you didn't specify how you wanted to _see_ it. Type `g + geom_point()` and see what happens.

```{r facet-xy, exercise = TRUE, exercise.lines = 5}

```
```{r facet-xy-hint}
g + #something goes here... a geom, perhaps?
```
```{r facet-xy-solution}
g + geom_point()
```

By calling the function `geom_point`, you added a visual **layer.** By _not_ assigning the expression to a variable, you _displayed_ the plot. 

You didn't have to pass any arguments to the function `geom_point`. That's because the object `g` has all the data stored in it. (Remember you saw that when you ran `summary` on `g` before.) 

## `ggplot` Scatterplot with Smoothed Regression Line

Use the expression you just typed (`g + geom_point()`) and add to it another layer, a call to `geom_smooth()`. Notice the message R gives you.


```{r geom-point-smooth, exercise = TRUE, exercise.lines = 5}

```
```{r geom-point-smooth-hint-1}
g + # What?
```
```{r geom-point-smooth-hint-2}
g + geom_point() + # what?
```
```{r geom-point-smooth-solution}
g + geom_point() + geom_smooth()
```

```{r what-method, echo=FALSE}
question("What method did `geom_smooth` use to calculate the regression line?",
         answer("loess", correct = TRUE),
         answer("none", message = "Does it have a method?"),
         answer("x ~ y", message = "That's not the method; it's something else."),
         answer("linear model", message = "That's not anywhere in the message from `ggplot`."),
         random_answer_order = TRUE,
         allow_retry = TRUE,
         correct = "That's correct!",
         incorrect = "Read the output from the last plot you made."
)
```

The gray shadow around the blue line is the _confidence band_. See how wide it is at the right?

Let's try a different smoothing function. Copy the expression you just typed, and instead of calling `geom_smooth` with no arguments, call it with the argument `method` set equal to the string "lm".

```{r geom-smooth-method, exercise = TRUE, exercise.lines = 5}

```
```{r geom-smooth-method-hint-1}
g + ?
```
```{r geom-smooth-method-hint-2}
g + geom_point() + ?
```
```{r geom-smooth-method-hint-3}
g + geom_point() + geom_smooth(...)
```
```{r geom-smooth-method-hint-4}
g + geom_point() + geom_smooth(method = "")
```
```{r geom-smooth-method-solution}
g + geom_point() + geom_smooth(method = "lm")
```

Changing the smoothing function to "lm" (linear model) `ggplot2` generated a linear regression line through the data. 

## `ggplot` Scatterplots in Facets, with Regression Lines

Copy the expression you just used and add to it another call, this time to the function `facet_grid`. Use the formula `. ~ drv` as it argument. Note that this is the same type of formula used in the calls to `qplot`. 

```{r face-grid-drv, exercise = TRUE, exercise.lines = 5}

```
```{r face-grid-drv-hint-1}
g + ?
```
```{r face-grid-drv-hint-2}
g + geom_point() + ?
```
```{r face-grid-drv-hint-3}
g + geom_point() + geom_smooth(method = "lm") + ?
```
```{r face-grid-drv-hint-4}
g + geom_point() + geom_smooth(method = "lm") + facet_grid(...)
```
```{r face-grid-drv-solution}
g + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ drv)
```

Each panel is labeled with the appropriate factor. 

All the data associated with 4-wheel drive cars is in the leftmost panel, front-wheel drive data is shown in the middle panel, and rear-wheel drive data in the rightmost. 

This is similar to the plot you created at the start of the lesson using `qplot`, although we used a different smoothing function than previously.

Have you noticed how we are building this plot function call by function call? That is part of the magic of `ggplot`: the way you can add layer after layer until you have the perfect plot for your needs.

## `ggplot` Scatterplot in Facets, with Regresssion Lines, with Annotations

So far you've just used the default labels that `ggplot` provides. You can add your own annotation using functions such as `xlab`, `ylab`, and `ggtitle`. 

In addition, the function `labs` is more general and can be used to label either or both axes in addition to providing an overall title. 

Copy the expression you just typed and add a call to the function `ggtitle` with the string argument "Swirl Rules!".

```{r ggtitle-first-use, exercise = TRUE, exercise.lines = 5}

```
```{r ggtitle-first-use-hint-1}
g + geom_point() + ?
```
```{r ggtitle-first-use-hint-2}
g + geom_point() + geom_smooth(method = "lm") + ?
```
```{r ggtitle-first-use-hint-3}
g + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ drv) +  ?
```
```{r ggtitle-first-use-hint-4}
g + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ drv) +  ggtitle(...)
```
```{r ggtitle-first-use-solution}
g + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ drv) +  ggtitle("Swirl Rules!")
```


## `ggplot` Customizations

Now that you've seen the basics, we'll talk about customizing. Each of the `geom` functions (e.g., `_point` and `_smooth`) has options to modify it. Also, the function `theme` can be used to modify aspects of the entire plot (e.g., the position of the legend). 

Two standard appearance themes are included in `ggplot`. These are `theme_gray`, which is the default theme (gray background with white grid lines), and `theme_bw`, which is a plainer (black and white) color scheme.

```{r customizations-intro, echo=FALSE}
question("Which of the following are customization options you can use with `ggplot`?",
         answer("`geom_point`", correct = TRUE),
         answer("`geom_smooth`", correct = TRUE),
         answer("`theme_bw`", correct = TRUE),
         answer("`theme_gray`", correct = TRUE),
         random_answer_order = TRUE,
         allow_retry = TRUE,
         correct = "That's right!",
         incorrect = "Remember how amazing the `ggplot2` package is? Select more options."
)
```

## Customizing Color, Size, and Alpha

Let's practice modifying aesthetics now. We'll use the graphic object `g` that we already filled with `mpg` data and add a call to the function `geom_point`, but this time we'll give `geom_point` 3 arguments. Set the argument `color` equal to "pink", the argument `size` to 4, and the argument `alpha` to 1/2. You'll see what `alpha` does when you run the plot.

All three of these aesthetic arguments are set equal to _constants_ _inside_ calls to `geom`s. We are not setting the aesthetics for the entire plot, but rather for the objects under the control of the `geom` call. And because they are constants, they can be passed directly to the `geom` function without being passed first to `aes`. More about that later.

Go ahead and create the plot.

```{r geom-point-mods, exercise = TRUE, exercise.lines = 5}

```
```{r geom-point-mods-hint-1}
g + ?
```
```{r geom-point-mods-hint-2}
g + geom_point(...)
```
```{r geom-point-mods-hint-3}
g + geom_point(color = "pink", ...)
```
```{r geom-point-mods-hint-4}
g + geom_point(color = "pink", size = 4, ...)
```
```{r geom-point-mods-solution}
g + geom_point(color = "pink", size = 4, alpha = 1/2)
```

See the different shades of pink? That's the result of the `alpha` aesthetic, which you set to 1/2. This aesthetic tells `ggplot` how transparent the points should be. 

```{r why-transparency, echo=FALSE}
question("Why would you want transparency?",
         answer("Because lighter circles are easier on the eyes.", message  = "That may be true, but it's not the reason we use `alpha`."),
         answer("Because reasons.", message = "You're not thinking. More tea?"),
         answer("Because darker circles then indicate values hit by multiple data points.",
                correct = TRUE),
         answer("Because variety is the spice of life.", message = "Yes, it is, but we are being more practical than fanciful at the moment."),
         random_answer_order = TRUE,
         allow_retry = TRUE,
         correct = "You are doing so well!",
         incorrect = ""
)
``` 

## Data-Dependent Customizations: Color, Size, and Alpha

Next, we'll modify the aesthetics so that `color` indicates which `drv` type each point represents. 

Again, use `g` and add to it a call to the function `geom_point` with 3 arguments. The first is `size` set equal to 4, the second is `alpha` equal to 1/2. The third is a call to the function `aes` with the argument `color` set equal to `drv`. 

Note that you **must** use the function `aes`: the color of the points depends on the data rather than being a constant (as it was in the previous example). 

```{r geom-point-color-drv, exercise = TRUE, exercise.lines = 5}

```
```{r geom-point-color-drv-hint-1}
g + ?
```
```{r geom-point-color-drv-hint-2}
g + geom_point(...)
```
```{r geom-point-color-drv-hint-3}
g + geom_point(aes(color = drv), ...)
```
```{r geom-point-color-drv-hint-4}
g + geom_point(aes(color = drv), size = 4, ...)
```
```{r geom-point-color-drv-solution}
g + geom_point(aes(color = drv), size = 4, alpha = 1/2)
```

Notice the helpful legend on the right decoding the relationship between `color` and `drv`.

```{r why-aes-this-time, echo=FALSE}
question("Why are some aesthetics passed to the `aes` function but others are not?",
         answer("Because programmers are a fickle breed", message = "Perhaps this is true, but it's not the reason for this particular design decision."),
         answer("Only constants must be passed to a `geom` by way of the `aes` function.", message = "That's backwards."),
         answer("Any aesthetic that is a constant can be passed directly to the `geom`; but an aesthetic that is an interpretation of the data you're plotting _must_ be passed to `geom` by way of the `aes` function.", correct = TRUE),
         answer("Don't know; don't care.", message = "Perhaps you need a cup of tea?"),
         random_answer_order = TRUE,
         allow_retry = TRUE,
         incorrect = "",
         correct = "Well done!"
)
```

## Customizing `labs` for Better Labels


Let's practice modifying labels. Again, we'll use `g` and add to it calls to 3 functions.

First, add a call to `geom_point` with an argument making the `color` dependent on the `drv` type (as we did in the previous example). Second, add a call to the function `labs` with the argument `title` set equal to "Swirl Rules!". Finally, add a call to `labs` with 2 arguments, one setting `x` equal to "Displacement" and the other setting `y` equal to "Hwy Mileage". 

```{r modify-labels, exercise = TRUE, exercise.lines = 5}

```
```{r modify-labels-hint-1}
g + geom_point(aes(...)) + ?
```
```{r modify-labels-hint-2}
g + geom_point(aes(color = drv)) + labs(...) + ?
```
```{r modify-labels-hint-3}
g + geom_point(aes(color = drv)) + labs(title = "Swirl Rules!") + labs(x = "?", ...)
```
```{r modify-labels-hint-4}
g + geom_point(aes(color = drv)) + labs(title = "Swirl Rules!") + labs(x = "Displacement", y = "?")
```
```{r modify-labels-solution}
g + geom_point(aes(color = drv)) + 
  labs(title = "Swirl Rules!") + 
  labs(x = "Displacement", 
       y = "Hwy Mileage")
```

If you haven't already, look at the hints in the code block above and note the judicious use of newlines to make the code more readable. This is good coding practice, as is adding the spaces around operators like `=` and after commas. Remember that you type code only once, but Future You (and others) might read your code over and over. Be considerate. Future You will need all possible help remembering what you wrote---and you won't be able to email Past You for an explanation. So make your code clear, and make it readable. 

```{r tf-labs, echo=FALSE}
question("True or False: You could have combined the two calls to the function `labs` in the previous example.",
         answer("True", correct = TRUE),
         answer("False"),
         random_answer_order = TRUE,
         allow_retry = TRUE
)
``` 

## Customizing `geom_smooth`

Let's practice customizing the `geom_smooth` calls. 

Use `g` and add to it a call to `geom_point`, setting the `color` to `drv` type (remember to use the call to the `aes` function because the value is data-dependent), `size` set to 2, and `alpha` set to 1/2. 

Then add a call to `geom_smooth` with 4 arguments. Set `size` equal to 4, `linetype` to 3, `method` to "lm", and `se` to `FALSE`.

```{r geom-smooth-customizations, exercise = TRUE, exercise.lines = 5}

```
```{r geom-smooth-customizations-hint-1}
g + geom_point(...) + ?
```
```{r geom-smooth-customizations-hint-2}
g + geom_point(aes(color = drv), ...) + 
  geom_smooth(...)
```
```{r geom-smooth-customizations-hint-3}
g + geom_point(aes(color = drv), 
               size = 2, alpha = 1/2) + 
  geom_smooth(size = 4, ...)
```
```{r geom-smooth-customizations-hint-4}
g + geom_point(aes(color = drv), 
               size = 2, 
               alpha = 1/2) +
  geom_smooth(size = 4, 
              ...)
```
```{r geom-smooth-customizations-solution}
g + geom_point(aes(color = drv), 
               size = 2, 
               alpha = 1/2) + 
  geom_smooth(size = 4,
              linetype = 3,
              method = "lm", 
              se = FALSE)
```


What did these arguments do? 

* The `method` specified a linear regression (note the negative slope indicating that the bigger the displacement the lower the gas mileage)    
* the `linetype` specified that the regression line should be dashed, not continuous    
* `size` made the dashes big    
* the `se` flag told `ggplot` to turn off the gray shadows we've been seeing around our regression lines that indicate standard errors (confidence intervals).

## Customizing by Adding a Theme

Finally, let's do a simple plot using the black and white theme, `theme_bw`. Specify `g` and add a call to the function `geom_point` with the argument setting the `color` to the `drv` type (inside a call to `aes`), `size` to 4, and `alpha` to 1/2. Then add a call to the function `theme_bw` with the argument `base_family` set equal to "Times". 

Are there differences? 

```{r fun-with-themes, exercise = TRUE, exercise.lines = 5}

```
```{r fun-with-themes-hint-1}
g + geom_point(...) + ?
```
```{r fun-with-themes-hint-2}
g + geom_point(aes(color = drv), ...) + ?
```
```{r fun-with-themes-hint-3}
g + geom_point(aes(color = drv), size = 4, alpha = 1/2) + theme_bw(...)
```
```{r fun-with-themes-hint-4}
g + geom_point(aes(color = drv), size = 4, alpha = 1/2) + theme_bw(base_family = "?")
```
```{r fun-with-themes-solution}
g + geom_point(aes(color = drv), size = 4, alpha = 1/2) + theme_bw(base_family = "Times")
```

```{r what-is-diff, echo=FALSE}
question("What is different now that you are using the black and white them?",
         answer("There are new gridlines"),
         answer("Everything is smaller"),
         answer("The gray background is gone", correct = TRUE),
         answer("The font type is different", correct = TRUE),
         random_answer_order = TRUE,
         allow_retry = TRUE,
         incorrect = "There are two things that are different, and one of them requires you to have good eyesight.",
         correct = "You got it!"
)
```

## Customizing Axis Limits

One final note before we go through a more complicated, layered `ggplot` example, and this concerns the limits of the axes. We're pointing this out to emphasize a subtle difference between `ggplot` and the `base` plotting function `plot`.

We've created some random `x` and `y` data, called `myx` and `myy`, components of a dataframe called `testdat`. These represent 100 random normal points, except halfway through, we made one of the points be an outlier. That is, we set its `y`-value to be out of range of the other points. 

Use the base plotting function `plot` to create a line plot of this data. Call it with 4 arguments: `myx`, `myy`, `type = "l"`, and `ylim = c(-3, 3)`. `type = "l"` tells `plot` you want to display the data as a line instead of as a scatterplot.

```{r myx-myy, exercise = TRUE, exercise.lines = 5}

```
```{r myx-myy-hint-1}
plot(...)
```
```{r myx-myy-hint-2}
plot(myx, myy, ...)
```
```{r myx-myy-hint-3}
plot(myx, myy, type = "l", ...)
```
```{r myx-myy-hint-4}
plot(myx, myy, type = "l", ylim = c(...))
```
```{r myx-myy-solution}
plot(myx, myy, type = "l", ylim = c(-3, 3))
```

`plot` plotted the points in the (-3, 3) range for y-values. The outlier is NOT shown on the line plot. 

We'll plot the same data with `ggplot`. Recall that the name of the dataframe is `testdat`. Create the graphical object `o` with a call to `ggplot` with 2 arguments: `testdat` (the data) and a call to `aes` with 2 arguments, `x` set equal to `myx`, and `y` set equal to `myy`.

```{r myxmyy-ggplot, exercise = TRUE, exercise.lines = 5}

```
```{r myxmyy-ggplot-hint-1}
o <- ? 
```
```{r myxmyy-ggplot-hint-2}
o <- ggplot(...) 
```
```{r myxmyy-ggplot-hint-3}
o <- ggplot(testdat, aes(...)) 
```
```{r myxmyy-ggplot-solution}
o <- ggplot(testdat, aes(x = myx, y = myy)) 
```

Nothing happened because we sent the output to the variable `o` rather than to your screen. 

Add a call to `geom_line` to `o` with 0 arguments.

```{r geomgline, exercise = TRUE, exercise.lines = 5}

```
```{r geomgline-hint}
o + ?
```
```{r geomgline-solution}
o + geom_line()
```

`ggplot` _did_ display the outlier point at (50, 100). As a result, the rest of the data is smashed down so you don't get to see what the bulk of it looks like. The single outlier probably isn't important enough to dominate the graph. How do we get `ggplot` to behave more like `plot` in a situation like this?

Let's take a guess that in addition to adding `geom_line` to `o` we also just have to add `ylim(-3, 3)` to it as we did with the call to `plot`. Try this now to see what happens.

```{r addylim-to-ggplot, exercise = TRUE, exercise.lines = 5}

```
```{r addylim-to-ggplot-hint-1}
o + ?
```
```{r addylim-to-ggplot-hint-2}
o + geom_line() + ?
```
```{r addylim-to-ggplot-hint-3}
o + geom_line() + ylim(...)
```
```{r addylim-to-ggplot-solution}
o + geom_line() + ylim(-3, 3)
```


`ggplot` simply ignored the outlier point at (50, 100). There's a break in the line which isn't very noticeable. 

Recall that at the beginning of the lesson we mentioned 7 components of a `ggplot` plot, one of which was a **coordinate system**. This is a situation where using a coordinate system would be helpful. 

Instead of adding `ylim(-3, 3)` to the expression `o + geom_line()`, add a call to the function `coord_cartesian` with the argument `ylim` set equal to `c(-3, 3)`.

```{r coord-cartesian, exercise = TRUE, exercise.lines = 5}

```
```{r coord-cartesian-hint-1}
o + ?
```
```{r coord-cartesian-hint-2}
o + geom_line() + ?
```
```{r coord-cartesian-hint-3}
o + geom_line() + coord_cartesian(...)
```
```{r coord-cartesian-hint-4}
o + geom_line() + coord_cartesian(ylim = c(...))
```
```{r coord-cartesian-solution}
o + geom_line() + coord_cartesian(ylim = c(-3, 3))
```

See the difference? This looks more like the plot produced by the base plot function. The outlier `y` value is not shown, but the plot indicates that it is larger than 3. 

Use any of these methods depending on your needs. But first, let's make the plot square by starting with `o`, adding `geom_line()`, adding `ylim(-3, 3)`, and then adding one more function with 2 arguments. See if you can guess what that final function should be. 

```{r one-more-function, exercise = TRUE, exercise.lines = 5}

```
```{r one-more-function-hint-1}
o + ?
```
```{r one-more-function-hint-2}
o + geom_line() + ?
```
```{r one-more-function-hint-3}
o + geom_line() + ylim(...) + ?
```
```{r one-more-function-hint-4}
o + geom_line() + ylim(-3, 3) + ?
```
```{r one-more-function-solution}
o + geom_line() + ylim(-3, 3) + xlim(-3, 3)
```

Now you know how to set limits along both axes. 

Also, you probably noticed that `ggplot`, unlike `plot`, ordered the data along the x-axis. We didn't ask it to do that; it's just nice that way. 

## A Complicated `ggplot` Example

We'll close with a complicated example to show you the full power of `ggplot` and the entire `ggplot2` package. We'll continue to work with the `mpg` dataset.

Start by creating the graphical object `ab` by assigning to it a call to `ggplot` with 2 arguments. The first is the dataset and the second is a call to the function `aes`. This call will have 3 arguments, `x` set equal to `displ`, `y` set equal to `hwy`, and `color` set equal to `factor(year)`. This last will allow us to distinguish between the two manufacturing years (1999 and 2008) in our data.

```{r complex-plot, exercise = TRUE, exercise.lines = 5}

```
```{r complex-plot-hint-1}
ab <- ggplot(...)
```
```{r complex-plot-hint-2}
ab <- ggplot(mpg, aes(...))
```
```{r complex-plot-hint-3}
ab <- ggplot(mpg, aes(x = displ, ...))
```
```{r complex-plot-hint-4}
ab <- ggplot(mpg, aes(x = displ, y = hwy, ...))
```
```{r complex-plot-solution}
ab <- ggplot(mpg, aes(x = displ, y = hwy, color = factor(year)))
```

Uh oh! Nothing happened. Does `ab` exist? Of course, it just isn't visible yet since you didn't add a visual layer. All you did was assign the data to the plot.

```{r if-you-typed-g, echo=FALSE}
question("If you typed `ab` at the command line, what would happen?",
         answer("a scatterplot would appear with 2 colors of points", 
                message = "Remember, you have not assigned a visual layer. How can there be color if there are no visuals?"),
         answer("R would return an error", correct = TRUE),
         answer("I would have to try this to answer the question", 
                message = "We'd like you to guess."),
         random_answer_order = TRUE,
         allow_retry = TRUE,
         correct = "Right!",
         incorrect = ""
)
```

We'll build the plot up step by step. First, add to `ab` a call to the function `geom_point` with 0 arguments.

```{r build-step-by-step, exercise = TRUE, exercise.lines = 5}

```
```{r build-step-by-step-hint}
ab + ?
```
```{r build-step-by-step-solution}
ab + geom_point()
```

A simple, yet comfortingly familiar scatterplot appears. Let's make our display a 2-dimensional multi-panel plot. Copy your last command and add to it a call the function `facet_grid`. Give it 2 arguments: the first is the formula `drv ~ cyl`, and the second is the argument `margins` set equal to `TRUE`. Try this now.

```{r familiar, exercise = TRUE, exercise.lines = 5}

```
```{r familiar-hint-1}
ab + ?
```
```{r familiar-hint-2}
ab + geom_point() + ?
```
```{r familiar-hint-3}
ab + geom_point() + facet_grid(...)
```
```{r familiar-hint-4}
ab + geom_point() + facet_grid(drv ~ cyl, ...)
```
```{r familiar-solution}
ab + geom_point() + facet_grid(drv ~ cyl, margins = TRUE)
```

A 4 by 5 plot, huh? The `margins` argument tells `ggplot` to display the marginal totals over each row and column, so instead of seeing 3 rows (the number of `drv` factors) and 4 columns (the number of `cyl` factors) we see a 3 + 1 = 4 by 4 + 1 = 5 display: one extra column for the row totals and one extra row for the column totals. 

Note that the panel in position (4, 5) is a tiny version of the scatterplot of the entire dataset. It contains the grand totals of data in rows by the grand total of data in columns.


Next, add to your last command (or retype it if you like to type) a call to `geom_smooth` with 4 arguments. These are `method` set to "lm", `se` set to `FALSE`, `size` set to 2, and `color` set to "black".

```{r add-or-retype, exercise = TRUE, exercise.lines = 10}

```
```{r add-or-retype-hint-1}
ab + geom_point() +
    ?
```
```{r add-or-retype-hint-2}
ab + geom_point() + 
  facet_grid(...) + 
  ?
```
```{r add-or-retype-hint-3}
ab + geom_point() + 
  facet_grid(drv ~ cyl, margins = TRUE) + 
  ?
```
```{r add-or-retype-hint-4}
ab + geom_point() + 
  facet_grid(drv ~ cyl, margins = TRUE) + 
  geom_smooth(...)
```
```{r add-or-retype-solution}
ab + geom_point() + 
  facet_grid(drv ~ cyl, margins = TRUE) + 
  geom_smooth(method = "lm", size = 2, se = FALSE, color = "black")
```

Angry Birds? 

Finally, add to your last command (or retype it if you like to type) a call to the function `labs` with 3 arguments. These are `x` set to "Displacement", `y` set to "Highway Mileage", and `title` set to "Swirl Rules!".

```{r finally, exercise = TRUE, exercise.lines = 5}

```
```{r finally-hint-1}
ab + geom_point() + ?
```
```{r finally-hint-2}
ab + geom_point() +
  facet_grid(drv ~ cyl, margins = TRUE) + 
  ?
```
```{r finally-hint-3}
ab + geom_point() +
  facet_grid(drv ~ cyl, margins = TRUE) + 
  geom_smooth(method = "lm", size = 2, se = FALSE, color = "black") + 
  ?
```
```{r finally-hint-4}
ab + geom_point() +
  facet_grid(drv ~ cyl, margins = TRUE) + 
  geom_smooth(method = "lm", size = 2, se = FALSE, color = "black") + 
  labs(...)
```
```{r finally-solution}
ab + geom_point() +
  facet_grid(drv ~ cyl, margins = TRUE) + 
  geom_smooth(method = "lm", size = 2, se = FALSE, color = "black") + 
  labs(x = "Displacement", y = "Highway Mileage", title = "Swirl Rules!")
```

You could have done these labels with separate calls to `labs`, but we thought you'd be sick of this by now. 

Anyway, congrats! You've concluded Part 2 of `ggplot2`. We hope you gotten enough mileage out of the lesson. If you like `ggplot2`, you can do some extras with the extra lesson.

Slides for this and other Data Science courses may be found at github https://github.com/DataScienceSpecialization/courses/. If you care to use them, they must be downloaded as a zip file and viewed locally. This lesson corresponds to 04_ExploratoryAnalysis/ggplot2. We relied on `swirl` lessons for the text. Thanks for making this great content open source!