---
title: "Paired _t_ Test"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
if(!require(ggplot2)){install.packages("ggplot2")}
if(!require(coin)){install.packages("coin")}
if(!require(pwr)){install.packages("pwr")}
library(learnr)
library(readr)
options(digits = 3, scipen = 9999)
crabs = read_csv("crabs.csv")
birds_Input = ("
 Bird   Typical  Odd
 A     -0.255   -0.324
 B     -0.213   -0.185
 C     -0.190   -0.299
 D     -0.185   -0.144
 E     -0.045   -0.027
 F     -0.025   -0.039
 G     -0.015   -0.264
 H      0.003   -0.077
 I      0.015   -0.017
 J      0.020   -0.169
 K      0.023   -0.096
 L      0.040   -0.330
 M      0.040   -0.346
 N      0.050   -0.191
 O      0.055   -0.128
 P      0.058   -0.182  
")

birds = read.table(textConnection(birds_Input),header=TRUE)
 
```

**Hint**: Use the Solution buttons to check your work or to get help if you can't figure out what the solution is.     
**Another Hint**: If you see a red ex next to your code, DO NOT RUN IT or this tutorial might freeze or crash. If it does freeze or crash, simply reload it. Your prior work will have been saved.

## The Purpose of this Lesson

This lesson is both a reference and a tutorial. It contains all the code samples you need to perform a paired _t_ test on your own data as well as background information such as how to interpret output and which graphs you might want to use to demonstrate results (and in some cases which graphs _not_ to use). 

If you already have a pretty good understanding of a paired _t_ test and just need a refresher, try just filling out the exercise code blocks and answering the questions. If you can do that without having to read the text, then there's really no reason to read the text except to sample our wit. Which you can do without, I'm sure. (Anyone for a good knock-knock joke?)

## What kinds of Variables work with a Paired _t_ Test

>* One [scalar (a.k.a. "measurement") variable](http://www.biostathandbook.com/variabletypes.html#measurement)     
* One [nominal variable](http://www.biostathandbook.com/variabletypes.html) with only two categories or two categories that you have selected. 

## Background

For a thorough treatment of the background for _t_ tests generally, see our lessons [One-Sample _t_ Test]("../sb-one-sample-t-test/") and [Two-Sample _t_ Test]("../sb-two-sample-t-test/").

Use the **paired t–test** when you have one measurement variable and two nominal variables, one of the nominal variables has only two values, and you have _only_ one observation for each combination of the nominal variables; in other words, you have pairs of observations rather than unrelated observations. 

The most common design is that one nominal variable represents individual organisms, while the other is "before" and "after" some treatment. Sometimes the pairs are spatial rather than temporal, such as left vs. right, injured limb vs. uninjured limb, etc. You can use the paired t–test for other pairs of observations; for example, you might sample an ecological measurement variable above and below a source of pollution in several streams.

The paired _t_ test shows whether the *mean difference* in the pairs is different from 0.

Paired t-tests can be conducted with the `t.test` function in the native `stats` package using the `paired = TRUE` option.  Data can be in long format or short format.

```{r a, echo=FALSE}
question("True or false: You can run a paired _t_ test using the same function you would use for a two-sample _t_ test.",
         answer("True", correct = TRUE),
         answer("False"),
         allow_retry = TRUE,
         random_answer_order = TRUE
)
```

```{r tfpair, echo=FALSE}
question("True or false: You can perform a paired _t_ test using the native `stats` package and data in either long or wide format.",
         answer("True", correct = TRUE),
         answer("False"),
         allow_retry = TRUE,
         random_answer_order = TRUE
)
```

## Null and Alternative Hypotheses

>* **H<sub>0</sub>**: "The mean difference between paired observations is zero. 

When the mean difference is zero, the means of the two groups must also be equal. Because of the paired design of the data, the null hypothesis of a paired t–test is usually expressed in terms of the **mean difference**.   

>* **H<sub>A</sub>**: "The mean difference between paired observations is not zero."    
* **H<sub>A</sub>** (1-sided): "The mean difference between paired observations is a positive number."        
* **H<sub>A</sub>** (1-sided): "The mean difference between paired observations is a negative number."


```{r bb, echo=FALSE}
question("What are NOT examples of a paired _t_ test null hypothesis?",
         answer("My electric bill is always high.", correct = TRUE),
         answer("People in my neighborhood pay just as much for electricity this year as last year.", correct = FALSE),
         answer("My electric bill is low no matter which year it is", correct = TRUE),
         answer("I watch a lot of television.", correct = TRUE),
         allow_retry = TRUE,
         random_answer_order = TRUE
)
```

## Assumptions

The paired t–test assumes that the _differences_ between pairs are normally distributed; you can use a histogram to check the normality. If the differences between pairs are severely non-normal, it would be better to use the [Wilcoxon signed-rank test](http://www.biostathandbook.com/wilcoxonsignedrank.html). 

We don't think the paired _t_ test is very sensitive to deviations from normality, so unless the deviation from normality is really obvious, you shouldn't worry about it.

The paired t–test does not assume that observations within each group are normal, only that the _differences_ are normal. And it does not assume that the groups are homoscedastic.

```{r whatisthenull, echo=FALSE}
question("The assumptions of a paired _t_ test are which?",
         answer("That the data is normally distributed"),
         answer("That the data is homeoscedastic"),
         answer("That the differences between the pairs are normally distributed", correct = TRUE),
         answer("That the differences between the pairs are homeoscedastic"),
         random_answer_order = TRUE,
         allow_retry = TRUE
)
```

## R Packages

Use the following code to install the packages used in this lesson:

```{r corregpackages, echo = TRUE}
if(!require(ggplot2)){install.packages("ggplot2")}
if(!require(coin)){install.packages("coin")}
if(!require(pwr)){install.packages("pwr")}
```

## How to run the Test

As mentioned above, you can do the test with wide or long data.

```{r wideorlongyn, echo=FALSE}
question("With which kind of data can you run a paired _t_ test?",
         answer("wide data", correct = TRUE),
         answer("long data", correct = TRUE),
         answer("pink data"),
         answer("data with a fox in a box"),
         random_answer_order = TRUE,
         allow_retry = TRUE
)
```

### Paired t-test, data in wide format, flicker feather example

Wiebe and Bortolotti (2002) examined color in the tail feathers of northern flickers. Some of the birds had one "odd" feather that was different in color or length from the rest of the tail feathers, presumably because it was regrown after being lost. They measured the yellowness of one odd feather on each of 16 birds and compared it with the yellowness of one typical feather from the same bird. 

There are two nominal variables, type of feather (typical or odd) and the individual bird, and one measurement variable, yellowness. Because these birds were from a hybrid zone between red-shafted flickers and yellow-shafted flickers, there was a lot of variation among birds in color, making a paired analysis more appropriate. 

I loaded a data set for you for this example called `birds`. Use `head(birds)` to see what the top of the data file looks like.

```{r headbirds, exercise = TRUE, exercise.lines = 5}

```
```{r headbirds-solution}
head(birds)
```

Pass the two variables to `t.test`, set `paired` equal to `TRUE`, and set `conf.level` to 0.95 as usual.

```{r pairdttestwideexample, exercise = TRUE, exercise.lines = 5}

```
```{r pairdttestwideexample-solution}
t.test(birds$Typical, 
       birds$Odd, 
       paired=TRUE, 
       conf.level=0.95)
```

The difference was significant with the odd feathers significantly less yellow than the typical feathers (higher numbers are more yellow).

```{r whatdoesthismean, echo=FALSE}
question("What do you conclude from the paired _t_ test you just ran?",
         answer("That the null hypothesis is true"),
         answer("That the true difference in means is not equal to 0", correct = TRUE),
         answer("That the difference between every pair of means is 0.137"),
         answer("That the null hypothesis can be safely rejected", correct = TRUE),
         random_answer_order = TRUE,
         allow_retry = TRUE
)
```

## Plotting the Results

Now let's make a simple plot of the differences. Follow these steps:

1. Create a variable called `difference` by subtracting `Typical` from `Odd` birds.

2. Send `difference` to `plot`, along with the parameter `pch` set to 16 and a `ylab` of "Difference (Odd - Typical)".

3. Call `abline` with 4 parameters: the first two are 0s, the third is `col` set to "blue", and the last is `lwd` set to 2.

4. Run your code!

```{r plotdiffs, exercise = TRUE, exercise.lines = 5}

```
```{r plotdiffs-solution}
difference = birds$Odd - birds$Typical
plot(difference,
     pch = 16,
     ylab="Difference (Odd – Typical)")

abline(0, 0, col="blue", lwd=2)
```

```{r bluelinemeaning, echo=FALSE}
question("How do you interpret the plot you just made?",
         answer("I don't interpret it. I just like looking at it."),
         answer("I see a relationship between Odd and Typical birds."),
         answer("A regression line would be negative."),
         answer("If the true differences of the means were 0, all the dots would be along the blue line", correct = TRUE),
         random_answer_order = TRUE,
         allow_retry = TRUE
)
```

Let's make another useful plot, a simple 1-to-1 plot of values. In this plot, each bird has its yellowness value of typical feathers along the x-axis and its value for yellowness of odd feathers along the y-axis. 

1. Send your two variables to `plot` and set `pch` to 16 as you did above. 

2. Also within the call to `plot`, set the parameter `xlab` to "Typical Feathers" and the parameter `ylab` to "Odd Feathers`. That's all you need to send to `plot`. s

3. On a new line, call the function `abline` with 4 parameters: 0, 1, `col` set equal to "blue", and `lwd` set equal to 2. 

```{r anotherusefulplot, exercise = TRUE, exercise.lines = 5}

```
```{r anotherusefulplot-solution}
plot(birds$Typical, birds$Odd,
     pch = 16,
     xlab="Typical feathers",
     ylab="Odd feathers")

abline(0,1, col="blue", lwd=2)
```

In this plot, circles below or to the right of the blue line indicate observations with a higher yellowness value for Typical than for Odd.

```{r bluelineslanted, echo=FALSE}
question("What does the blue line represent in the plot you just made?",
         answer("Where the dots would be the differences between the yellowness values was 0", correct = TRUE),
         answer("A positive regression line"),
         answer("The line above which values are illegal"),
         random_answer_order = TRUE,
         allow_retry = TRUE
)
```

## Note to the Learner

This lesson is not yet complete. We hope you enjoyed the first part of it. Stay tuned for more to come! 