---
title: "A Simple Aggregation Exercise: Piping and Summarizing"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
options(digits = 3, scipen = 9999)
if(!require(learnr)){install.packages("learnr")}
library(learnr)
this_Input =("
Group Value
2pm    69    
2pm    70    
2pm    66    
2pm    63    
2pm    68    
2pm    70    
2pm    69    
2pm    67    
2pm    62    
2pm    63    
2pm    76    
2pm    59    
2pm    62    
2pm    62    
2pm    75    
2pm    62    
2pm    72     
2pm    63    
5pm    68
5pm    62
5pm    67
5pm    68
5pm    69
5pm    67
5pm    61
5pm    59
5pm    62
5pm    61
5pm    69
5pm    66
5pm    62
5pm    62
5pm    61
5pm    70
")
ht = read.table(textConnection(this_Input),header=TRUE)
names(ht) <- c("section", "inches")
tt_height <- t.test(inches ~ section, 
                    data =ht,
                    var.equal = TRUE,
                    conf.level = 0.95)
p_height <- round(tt_height$p.value, 2)
```

Let's pipe and summarize using simple example data. It's already loaded for you: a dataset with two variables, `section` and `inches`, that contains the heights of undergraduates in two of John McDonald's Biostatics sections, the 2 pm one and the 5 pm one.  

First, find out what the means are for the two groups. We'll group the data and summarize it. Make sure you have `dplyr` in the library. It's a very good data manipulation package. Also put `magrittr` in the library so we can use `%>%` for piping ([What's piping?](https://www.datacamp.com/community/tutorials/pipe-r-tutorial) Unless you're in interested in the history of R, control- or command-f the phrase "Why use it?" and start reading from there).

Group the data by `section` using `group_by`. Let's also using the `magrittr` package's piping (`%>%`) for the commands so we can stack them on each other one by one. 

We will derive a new dataset from `ht` called `by_section`. Assign to it `ht`, then add  `%>%` to pipe `ht` to a few commands. The first one is `dplyr::group_by` (we're specifying the package `dplyr` to call `group_by` to avoid confusion in case we have any other packages loaded with similar commands that do different things. This is good coding practice and I don't do it enough). Since `ht` is at the start of the pipeline, you don't have to tell `group_by` which data you're using. Just pass it the grouping variable, here `section`. Thats the end of that command. Add also a command to have a look at `by_section`, using whatever method you like (just type the name `by_section` or use `head`).

```{r group_by, exercise = TRUE}



```
```{r group_by-solution}
library(dplyr)
library(magrittr)
class(ht$inches)
by_section <- ht %>% 
   dplyr::group_by(section) %>%
   dplyr::summarise(
      ht_sd = sd(inches),
      ht_mean = mean(inches),
      ht_n = n()
   )
by_section
```

Now revise the code above, adding a `%>%` on the same line, after the `group_by` command. On the next line, call `dplyr::summarise`, passing it a three arguments: `ht_sd = sd(inches)` to get the standard deviation of `inches`, `ht_mean = mean(inches)` to get the mean of `inches`, and `ht_n = n()` to get the count in each section. Execute the code chunk, not deleting whatever method you used to look at `by_section` as a separate command (i.e., don't pipe to it). 

You've just aggregated data using the `dplyr` and `magrittr` packages. Well done!

