---
title: "A Simple Aggregation Exercise: Piping and Summarizing"
output: 
 learnr::tutorial:
    progressive: true
    allow_skip: true
    css:
      - www/bootstrap.min.css
      - www/flexdashboard.min.css
      - www/style.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
options(digits = 3, scipen = 9999)
if(!require(learnr)){install.packages("learnr")}
library(learnr)
library(dplyr)
tutorial_options(exercise.startover = TRUE,
   exercise.eval = TRUE, 
   exercise.cap = "R Code",
   exercise.completion = TRUE,
   exercise.diagnostics = TRUE)
   
this_Input =("
Group Value
2pm    69    
2pm    70    
2pm    66    
2pm    63    
2pm    68    
2pm    70    
2pm    69    
2pm    67    
2pm    62    
2pm    63    
2pm    76    
2pm    59    
2pm    62    
2pm    62    
2pm    75    
2pm    62    
2pm    72     
2pm    63    
5pm    68
5pm    62
5pm    67
5pm    68
5pm    69
5pm    67
5pm    61
5pm    59
5pm    62
5pm    61
5pm    69
5pm    66
5pm    62
5pm    62
5pm    61
5pm    70
")
ht = read.table(textConnection(this_Input),header=TRUE)
names(ht) <- c("section", "inches")
tt_height <- t.test(inches ~ section, 
                    data =ht,
                    var.equal = TRUE,
                    conf.level = 0.95)
p_height <- round(tt_height$p.value, 2)
```

## Piping and Summarizing with `dplyr`

Let's pipe and summarize using simple example data. It's already loaded for you: a dataset called `ht` with two variables, `section` and `inches`, that contains the heights of undergraduates in two of John McDonald's Biostatics sections, the 2 pm one and the 5 pm one. Use `head(ht)` to see the top of this data file.

```{r seetopht, exercise = TRUE, exercise.lines = 5}

```
```{r seetopht-solution}
head(ht)
```

What classes are the two variables in `ht`?

```{r whatclasseshtvars, exercise = TRUE, exercise.lines = 5}

```
```{r whatclasseshtvars-solution}
class(ht$section)
class(ht$inches)
```

`section` is a factor. Factors are excellent grouping variables. `inches` is an integer, which means we can perform mathematical functions such as `mean` or `sd` on it.

To find out more about why variable types are important to understand at the outset of any examination of your data, see our [Variables lesson](../variables/). 

Find out what the means are for the two groups. We'll group the data and summarize it. Make sure you have `dplyr` in the library. `dplyr` puts `magrittr` in the library so we can use `%>%` for piping ([What's piping?](../piping-and-summarizing/)) Unless you're in interested in the history of R, control- or command-f the phrase "Why use it?" and start reading from there).

Group the data by `section` using `group_by(section, .data = ht)` and examine the output.

```{r label, exercise = TRUE, exercise.lines = 5}

```
```{r label-solution}
library(dplyr)
group_by(section, .data = ht)
```

Not much to see, really, as you haven't `summarize`d anything. `group_by` organizes summary data. Let's create a pipeline that includes grouping _and_ summarizing. 


We will derive a new dataset from `ht` called `by_section`. Assign to it `ht`, then add  `%>%` to pipe `ht` to a few commands. The first one is `dplyr::group_by` (we're specifying the package `dplyr` to call `group_by` to avoid confusion in case we have any other packages loaded with similar commands that do different things. This is good coding practice and I don't do it enough). Since `ht` is at the start of the pipeline, you don't have to tell `group_by` which data you're using. Just pass it the grouping variable, here `section`. Thats the end of that command. Then add a command so we can see `by_section`, using whatever method you like (just type the name `by_section` or use `head`).

```{r group_by, exercise = TRUE}



```
```{r group_by-solution}
by_section <- ht %>% 
   dplyr::group_by(section) 
by_section
```

Now duplicate the code from above and add a `%>%` on the `group_by` line. On the next line, call `dplyr::summarise`, passing it a three arguments: `ht_sd = sd(inches)` to get the standard deviation of `inches`, `ht_mean = mean(inches)` to get the mean of `inches`, and `ht_n = n()` to get the count in each section. Execute the code chunk, not deleting whatever method you used to look at `by_section` as a separate command (i.e., don't pipe to it). 

```{r buildyourpipeline, exercise = TRUE, exercise.lines = 5}

```
```{r buildyourpipeline-solution}
by_section <- ht %>% 
   dplyr::group_by(section) %>%
   dplyr::summarise(
      ht_sd = sd(inches),
      ht_mean = mean(inches),
      ht_n = n()
   )
by_section
```

You've just aggregated data using the `dplyr` and `magrittr` packages. Well done!

